The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-01 src/ms.tex
Spectral coordinates

Summarize major pieces of development since the last paper (since v2.0). A few
ideas (taken from What's New pages) below, but feel free to expand on or remove
from this list:
\begin{itemize}
    \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
    \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
    \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
    \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
    \item APE 14 WCS
    \item A new SpectralCoord class for representing and transforming spectral quantities
    \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
\end{itemize}

>>>>

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with (purposefully) few changes to the
software interface.
This maturity and stability allows for the broader astronomy \python community
to rely upon the \astropypkg core package and build specialized packages within
the Astropy ecosystem.
Even many non-astronomical \python libraries have come to rely on \astropypkg.

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citealt{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

\subsubsection*{New Features} \label{sec:core-features-new}

\begin{itemize}
%  \item Time Series sub-package (V3.1: New Box Least Squares Periodogram)
%  \item Uncertainties and Distributions
  \item ERFA module is now it’s own thing; Obliquely tie in that units
    remains in astropy
  \item ? Pytest plugin ecosystem
%  \item A new SpectralCoord class for representing and transforming spectral quantities
%    Velocity in Coordinates
  \item WCS API : A shared Python interface for World Coordinate Systems
        Include APE reference
\end{itemize}

New features come in various forms, from new submodules providing
wholly new capabilities, to larger additions to existing submodules,
to large rewrites of the interface of modules.  This range is spanned
by the four main features that entered the astropy core, which we
describe below.

\paragraph{Uncertainties and Distributions.} Early in our carreers, we
were likely all taught that a measurement without a unit or an
uncertainty is essentially meaningless.  \astropyunits already
provides the ability to associate numbers with units, and propagate
these correctly, with its \astropyQuantity\ class, but so far the
ability to associate and propagate uncertainties is relatively
limited: in \astropysubpkg{nddata}, there are options to associate
errors with data arrays, but these are treated as independent, i.e.,
covariances are not tracked.

In general, proper error propagation is difficult, and often it is
best to use Monte Carle methods.  With the new
\astropysubpkg{uncertainties} sub-package, a first step is made to
make this possible.  It allows one to generate, for each variable,
randomly drawn samples in a \astropyDistribution, and then propagate
these by passing them through the normal analysis, producing a
\astropyDistribution\ of final results that can be inspected.

This \astropyDistribution\ can be any type of array, including of
course a \astropyQuantity, and the goal is to ensure it can be
seamlessly used to instantiate also other astropy classes, such as
\astropyskycoord\ and \astropyTime.

\paragraph{Time series.}  From sampling a continuous variable at fixed
times to counting events binned into time windows, many different
areas of astrophysics require the manipulation of time series.  These
are often logically tables, and the new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to tables of data as a
function of time, where the data can either represent samples or
averages over particular time bins.  The new classes offer a number of
special methods to manipulate time series (folding, resampling, etc.)
and to read different data formats (such as {\em Kepler} lightcurves,
etc.).

Also part of \astropysubpkg{timeseries} are common analysis routines,
including Lomb-Scargle and box-least-squares periodogrammes.

\paragraph{Spectral Coordinates.} Measurements are often taken at
specific ``spectral coordinates'', be they frequencies, wavelengths,
or photon energies.  The new \astropySpectralCoord\ class allows to
transform between these as well as to equivalent Doppler velocities.
It builds on the \astropyQuantity, which can also represent these
kinds of physical quantities, and allows conversion via dedicated
equivalencies, but aims to make this more straightforward.
Furthermore, it can be made aware of the observer and target reference
frames, allowing for instance transformation from telescope-centric (or
topocentric) frames to Barycentric or Local Standard of Rest
(LSRK and LSRD) velocity frames.

\subsubsection*{Interoperability} \label{sec:core-features-interoperability}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-precision_accuracy_repreducibility}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}
